## Description of the data

This section is divided into two subsections: one describing the clinical data and one describing the count data.

### Clinical data

First, we extract the counts for some variables in the dataset

For the gender variable in our dataset:

```{r}
clinical_data_augment |> 
  count(gender.demographic) |> 
  mutate(percentage = n *100/ sum(n))
```

As it can be seen, the dataset is quite balanced in this variable, with almost the same name of samples in terms of gender.

For the variable displaying the race of the patients from each sample:

```{r}
clinical_data_augment |> 
  count(race.demographic) |> 
  mutate(percentage = n *100 / sum(n))
```

It can be observed that the dataset is quite unbalanced in terms of race, as the vast majority of the samples (89%) come from white people.

For the cancer stages, we can observe in the table below that most of the samples come from patients of stage I (\~ 60%) between IA and IB.

```{r}
clinical_data_augment |> 
  count(ajcc_pathologic_stage.diagnoses) |>
  mutate(percentage = n*100 / sum(n))
```

For the age, as it is a numerical value we can calculate its mean, median, and standard deviation:

```{r}
clinical_data_augment |> 
  summarise(
    mean_age = mean(age_at_index.demographic, na.rm = TRUE), 
    median_age = median(age_at_index.demographic, na.rm = TRUE),
    sd_age = sd(age_at_index.demographic, na.rm = TRUE)
)
```

After calculating these descriptive statistics related to the age we observe a mean of 66.7 years old, a median of 68 years old and a standard deviation of 9.5 years.

We can also print the amount of samples from each of the smoker groups:

```{r}
clinical_data_augment |> 
  count(smoker_groups) |>
  mutate(percentage = n * 100/ sum(n))
```

46% of the samples in the dataset do not provide any meaningful information. The most predominant smoker group is the heavy smoker ( \>= 40 pack years smoked) with a \~33%.

We can extract the average age by smoker group:

```{r}
clinical_data_augment |>
  group_by(smoker_groups) |>
  summarize(mean_age = mean(age_at_index.demographic, 
                            na.rm = TRUE))

```

As it can be observed in the table above, the patients from the light smokers group have a lower mean age (61.57%) than the rest of the groups.

And the percentage of women and men alive and deceased:

```{r}
clinical_data_augment |>
  count(gender.demographic, vital_status.demographic) |>
  group_by(gender.demographic) |>
  mutate(percentage = n*100 / sum(n)) |>
  ungroup()
```

From this table we can observe that the majority of patients whose sample are in the study are still alive. This difference is especially predominant in women, where 77% of the samples are from patients who are still alive at the time of the study.

For each gender, we can also extract the number of samples corresponding to each cancer stage:

```{r}
clinical_data_augment |>
  count(gender.demographic, 
        ajcc_pathologic_stage.diagnoses) |>
  spread(ajcc_pathologic_stage.diagnoses, n, fill = 0)
```

As it can be seen, the samples are quite distributed among men and women, except for in Stage IV cancer, where all the samples except for 1 come from men.

Once we have displayed some counts for the relevant variables, we move on to plotting for a clearer visualization:

First, we will create the results folder if it has not been created. Here we will store the plots:

```{r}
dir.create(here("results"))
```

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=smoker_groups, 
                fill= smoker_groups)) +
  geom_bar() +
  labs(
    x = "Smoker groups",
    y = "Count of samples",
    title = "Samples per smoker group",
    fill="Smoker groups"
  )
  
```

```{r}
ggsave(here("results/04_samples_per_smoker_group.png"), plot= last_plot() )
```

Here we have a bar chart showing the count of samples grouped by smoking history. Heavy smokers (\>= 40 pack years) are the predominant group (\~33%) out of the smokers, while \~46% of samples lack meaningful smoking data.

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=gender.demographic, 
                fill= gender.demographic)) +
  geom_bar() +
  labs(
    x = "Gender",
    y = "Count of samples",
    title = "Samples per gender",
    fill="Gender"
  )
```

```{r}
ggsave(here("results/04_samples_per_gender.png"), plot= last_plot() )
```

This is a bar chart displaying the count of samples categorized by gender relatively balanced, with almost an equal number of samples for males and females.

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=ajcc_pathologic_stage.diagnoses, 
                fill= ajcc_pathologic_stage.diagnoses)) +
  geom_bar() +
  labs(
    x = "Cancer stage",
    y = "Count of samples",
    title = "Samples per cancer stage",
    fill= "Cancer stage"
  )
```

```{r}
ggsave(here("results/04_samples_cancer_stage.png"), plot= last_plot() )
```

This bar chart shows the distribution of samples across different cancer stages, representing \~60% of the samples from patients in Stage I.

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=age_at_index.demographic)) +
  geom_histogram(binwidth = 3, 
                 fill = "lightblue", color = "black") +
  labs(
    x = "Age",
    y = "Count of samples",
    title = "Distribution of ages"
  )
```

```{r}
ggsave(here("results/04_age_histogram.png"), plot= last_plot() )
```

This histogram illustrates the age distribution of the samples. The majority of patients are aged between 60 and 75, with a mean age of 66.7 and a median age of 68.

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=ajcc_pathologic_stage.diagnoses, 
                y=age_at_index.demographic, 
                fill= ajcc_pathologic_stage.diagnoses)) +
  geom_boxplot() +
  labs(
    x = "Cancer stage",
    y = "Age",
    title = "Age distribution by cancer stage",
    fill="Cancer stage"
  )
```

```{r}
ggsave(here("results/04_age_cancer_stage.png"), plot= last_plot() )
```

This boxplot compares the age distributions of patients across different cancer stages. Samples belonging to stage IV patients present a range of ages lower compared to prior stages.

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=vital_status.demographic, 
                y=age_at_index.demographic, 
                fill= vital_status.demographic)) +
  geom_boxplot() +
  labs(
    x = "Vital status",
    y = "Age at diagnosis",
    title = "Vital status vs age at diagnosis",
    fill= "Vital status"
  )
```

```{r}
ggsave(here("results/04_vital_status_vs_age.png"), plot= last_plot() )
```

Here we have a boxplot comparing the age at diagnosis of patients grouped by their vital status. The majority of patients in the dataset are alive.

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=ajcc_pathologic_stage.diagnoses, 
                fill= smoker_groups)) +
  geom_bar(position="fill") +
  labs(
    x = "Cancer stage",
    y="Proportion of smoker group",
    title = "Smoker groups within each cancer stage",
    fill="Smoker groups"
  )

```

```{r}
ggsave(here("results/04_smoker_groups_within_cancer_stage.png"), plot= last_plot() )
```

This is a stacked bar chart showing the proportion of smoker groups for each cancer stage. The most remarkable situation is that all the samples regarding stage IIIB belong to heavy smokers.

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=smoker_groups, 
                fill=ajcc_pathologic_stage.diagnoses)) +
  geom_bar(position="fill") +
  labs(
    x = "Smoker group",
    y="Proportion of cancer stages",
    title = "Cancer stage within each smoker group",
    fill="Cancer stage"
  )
```

```{r}
ggsave(here("results/04_cancer_stage_within_smoker_group.png"), plot= last_plot() )
```

Here is a stacked bar chart showing the proportion of cancer stages within each smoker group . In all the smoker groups the predominant stage of cancer is Stage IA and IB.

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=ajcc_pathologic_stage.diagnoses, 
                fill= gender.demographic)) +
  geom_bar(position="fill") +
  labs(
    x = "Cancer stage",
    y="Gender",
    title = "Gender distribution within cancer stages",
    fill="Gender"
  )

```

```{r}
ggsave(here("results/04_gender_within_cancer_stages.png"), plot= last_plot() )
```

Here is a stacked bar chart showing the distribution of genders within each cancer stage. In earlier stages of cancer there is a balance distribution between genders, while in later stages we have a predominance of male samples (Stage IV).

```{r}
clinical_data_augment |> 
  ggplot(
    aes(x= age_at_index.demographic,
        fill= ajcc_pathologic_stage.diagnoses)) +
      geom_histogram(binwidth= 15, 
                     position="dodge") +
      facet_wrap(~gender.demographic) +
      labs(title="Age distribution by cancer stage faceted by gender",
           fill= "Cancer stage",
           x= "Age",
           y= "Sample count")

```

```{r}
ggsave(here("results/04_age_by_cancer_stage_gender.png"), plot= last_plot() )
```

Here we have a faceted histogram showing the age distribution within each cancer stage, separated by gender. The distribution differ a lot between male and female samples.

```{r}
clinical_data_augment |>
  group_by(race.demographic) |>
  filter(n() >= 2) |> #here we keep only the races with more than 2 samples for a clearer representation
  ungroup() |>
  ggplot(aes(x = race.demographic, 
             y = age_at_index.demographic, 
             fill = race.demographic)) +
  geom_violin() +
  labs(title = "Age at diagnosis by race", 
       x = "Race", 
       y = "Age at diagnosis",
       fill="Race")


```

```{r}
ggsave(here("results/04_age_by_race_violin.png"), plot= last_plot() )
```

Here we have a violin plot comparing the age distribution at diagnosis for different races. For samples from Asians the distribution is narrower and this indicates that there is less variation in the age at diagnosis. Whites have a wider range of age at diagnosis.

```{r}
clinical_data_augment |>
  ggplot(aes(x = gender.demographic, 
             y = age_at_index.demographic, 
             fill = gender.demographic)) +
  geom_violin() +
  labs(title = "Age at diagnosis by gender", 
       x = "Gender", 
       y = "Age at Diagnosis",
       fill="Gender")
```

```{r}
ggsave(here("results/04_age_by_gender_violin.png"), plot= last_plot() )
```

This is a violin plot comparing the age distribution at diagnosis between genders. In males the diagnoses are mostly around one age group (70 years old) while for females the diagnosis are in two main age groups (60 and 70 years old) .

### Count data

Histogram displaying the frequency of the genes across the samples

```{r}
# We first convert our count data into long format
gene_presence <- count_data_clean |>
  pivot_longer(
    cols = -1, # We exclude the first column (samples)
    names_to = "gene",
    values_to = "count"
  ) |>
  filter(count > 0) |>
  group_by(gene) |>
  summarize(samples_present = n(), .groups = "drop")
```

```{r}
gene_presence |>
  ggplot(aes(x = samples_present)) +
  geom_histogram(binwidth = 5, 
                 fill = "lightblue", 
                 color = "black") +
  labs(
    title = "Distribution of gene presence",
    x = "Number of samples a gene is present in",
    y = "Frequency of genes"
  ) 
```

```{r}
ggsave(here("results/04_gene_presence_distribution.png"), plot= last_plot() )
```

This is a histogram displaying the distribution of gene presence across the samples. Most of the genes are present in a large number of samples.
