## Description of the data

This section is divided into two subsections: one describing the clinical data and one describing the count data.

### Clinical data

First, we extract the counts for some variables in the dataset

For the gender variable in our dataset:

```{r}
clinical_data_augment |> 
  count(gender.demographic) |> 
  mutate(percentage = n *100/ sum(n))
```

As it can be seen, the dataset is quite balanced in this variable, with almost the same name of samples in terms of gender.

For the variable displaying the race of the patients from each sample:

```{r}
clinical_data_augment |> 
  count(race.demographic) |> 
  mutate(percentage = n *100 / sum(n))
```

It can be observed that the dataset is quite unbalanced in terms of race, as the vast majority of the samples (89%) come from white people.

For the cancer stages, we can observe in the table below that most of the samples come from patients of stage I (\~ 60%) between IA and IB.

```{r}
clinical_data_augment |> 
  count(ajcc_pathologic_stage.diagnoses) |>
  mutate(percentage = n*100 / sum(n))
```

For the age, as it is a numerical value we can calculate its mean, median, and standard deviation:

```{r}
clinical_data_augment |> 
  summarise(
    mean_age = mean(age_at_index.demographic, na.rm = TRUE), 
    median_age = median(age_at_index.demographic, na.rm = TRUE),
    sd_age = sd(age_at_index.demographic, na.rm = TRUE)
)
```

After calculating these descriptive statistics related to the age we observe a mean of 66.7 years old, a median of 68 years old and a standard deviation of 9.5 years.

We can also print the amount of samples from each of the smoker groups:

```{r}
clinical_data_augment |> 
  count(smoker_groups) |>
  mutate(percentage = n * 100/ sum(n))
```

46% of the samples in the dataset do not provide any meaningful information. The most predominant smoker group is the heavy smoker ( \>= 40 pack years smoked) with a \~33%.

We can extract the average age by smoker group:

```{r}
clinical_data_augment |>
  group_by(smoker_groups) |>
  summarize(mean_age = mean(age_at_index.demographic, na.rm = TRUE))

```

As it can be observed in the table above, the patients from the light smokers group have a lower mean age (61.57%) than the rest of the groups.

And the percentage of women and men alive and deceased:

```{r}
clinical_data_augment |>
  count(gender.demographic, vital_status.demographic) |>
  group_by(gender.demographic) |>
  mutate(percentage = n*100 / sum(n)) |>
  ungroup()
```

From this table we can observe that the majority of patients whose sample are in the study are still alive. This difference is especially predominant in women, where 77% of the samples are from patients who are still alive at the time of the study.

For each gender, we can also extract the number of samples corresponding to each cancer stage:

```{r}
clinical_data_augment |>
  count(gender.demographic, ajcc_pathologic_stage.diagnoses) |>
  spread(ajcc_pathologic_stage.diagnoses, n, fill = 0)
```

As it can be seen, the samples are quite distributed among men and women, except for in Stage IV cancer, where all the samples except for 1 come from men.

Once we have displayed some counts for the relevant variables, we move on to plotting for a clearer visualization:

First, we will create the results folder if it has not been created. Here we will store the plots:

```{r}
dir.create(here("results"))
```

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=smoker_groups, fill= smoker_groups)) +
  geom_bar() +
  labs(
    x = "Smoker groups",
    y = "Count of samples",
    title = "Samples per smoker group",
    fill="Smoker groups"
  )
  
```

```{r}
ggsave(here("results/04_samples_per_smoker_group.png"), plot= last_plot() )
```

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=gender.demographic, fill= gender.demographic)) +
  geom_bar() +
  labs(
    x = "Gender",
    y = "Count of samples",
    title = "Samples per gender",
    fill="Gender"
  )
```

```{r}
ggsave(here("results/04_samples_per_gender.png"), plot= last_plot() )
```

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=ajcc_pathologic_stage.diagnoses, fill= ajcc_pathologic_stage.diagnoses)) +
  geom_bar() +
  labs(
    x = "Cancer stage",
    y = "Count of samples",
    title = "Samples per cancer stage",
    fill= "Cancer stage"
  )
```

```{r}
ggsave(here("results/04_samples_cancer_stage.png"), plot= last_plot() )
```

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=age_at_index.demographic)) +
  geom_histogram(binwidth = 3, fill = "lightblue", color = "black") +
  labs(
    x = "Age",
    y = "Count of samples",
    title = "Distribution of ages"
  )
```

```{r}
ggsave(here("results/04_age_histogram.png"), plot= last_plot() )
```

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=ajcc_pathologic_stage.diagnoses, y=age_at_index.demographic, fill= ajcc_pathologic_stage.diagnoses)) +
  geom_boxplot() +
  labs(
    x = "Cancer stage",
    y = "Age",
    title = "Age distribution by cancer stage"
  )
```

Vital status vs age at diagnosis

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=vital_status.demographic, y=age_at_index.demographic/365, fill= vital_status.demographic)) +
  geom_boxplot() +
  labs(
    x = "Vital status",
    y = "Age at diagnosis",
    title = "..."
  )
```

Cancer stage across smoker groups (stacked bar chart)

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=ajcc_pathologic_stage.diagnoses, fill= smoker_groups)) +
  geom_bar(position="fill") +
  labs(
    x = "Cancer stage",
    y="Proportion of smoker group",
    title = "Smoker groups within each cancer stage"
  )
```

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=smoker_groups, fill=ajcc_pathologic_stage.diagnoses)) +
  geom_bar(position="fill") +
  labs(
    x = "Smoker group",
    y="Proportion of cancer stages",
    title = "Cancer stage within each smoker group"
  )
```

cancer stage and gender:

```{r}
clinical_data_augment |>
  ggplot(
    mapping=aes(x=ajcc_pathologic_stage.diagnoses, fill= gender.demographic)) +
  geom_bar(position="fill") +
  labs(
    x = "Cancer stage",
    y="Gender",
    title = "Gender distribution within cancer stages"
  )
```

```{r}
clinical_data_augment |> 
  ggplot(
    aes(x= age_at_index.demographic,
        fill= ajcc_pathologic_stage.diagnoses)) +
      geom_histogram(binwidth=15,position="dodge") +
      facet_wrap(~gender.demographic) +
      labs(title="Age distribution by cancer stage, faceted by gender")
```

```{r}
clinical_data_augment |>
  group_by(race.demographic) |>
  filter(n() >= 2) |> #here we keep only the races with more than 2 samples for a clearer representation
  ungroup() |>
  ggplot(aes(x = race.demographic, y = age_at_index.demographic, fill = race.demographic)) +
  geom_violin() +
  labs(title = "Age at Diagnosis by Race", x = "Race", y = "Age at Diagnosis")

```

```{r}
clinical_data_augment |>
  ggplot(aes(x = gender.demographic, y = age_at_index.demographic, fill = gender.demographic)) +
  geom_violin() +
  labs(title = "Age at Diagnosis by gender", x = "gender", y = "Age at Diagnosis")
```

### Count data

Histogram displaying the frequency of the genes across the samples

```{r}
# Convert the data into a long format for tidyverse operations
gene_presence <- count_data_clean |>
  pivot_longer(
    cols = -1, # Exclude the first column (samples)
    names_to = "gene",
    values_to = "count"
  ) |>
  filter(count > 0) |>
  group_by(gene) |>
  summarize(samples_present = n(), .groups = "drop")
```

```{r}
gene_presence |>
  ggplot(aes(x = samples_present)) +
  geom_histogram(binwidth = 50) +
  labs(
    title = "Distribution of Gene Presence Across Samples",
    x = "Number of Samples a Gene is Present In",
    y = "Frequency of Genes"
  ) 
```
