---
title: "05_Analysis_01"
format: html
editor: visual
---

## Differential gene expresion analysis

Differential gene expression analysis will be conducted using the R package "DESeq2". The objective is to identify genes that are signicantly differentlly expressed among patients with lung adenocarcinoma across three distict groups: light, medium and heavy smokers.

```{r}

#Load libraries

library("DESeq2")
library("tidyverse")
library("ggrepel")
```

### Step 1: preparing data for analysis

```{r}

# Mutating smoker_groups variable to ensure it is treated as a factor

sample_table <-sample_table |>
  mutate(smoker_groups=factor(smoker_groups))

# Create a DESeq2 Dataset

dds <- DESeqDataSetFromMatrix(
  countData = count_data_clean_DEseq_trans, 
  colData = sample_table_DEseq,
  design= ~ smoker_groups
)


dds <- DESeq(dds)
resultsNames(dds) # lists the coefficients
```

```{r}

res <- results(
  dds, 
  contrast = list(c("smoker_groups_heavy_vs_light", "smoker_groups_medium_vs_light")),
  alpha=0.05
)

# or to shrink log fold changes association with condition:

summary(res)
```

Convert results from DESeq to a tidy dataframe, select "gene", "log2FoldChange" and "pvalue" columns.

```{r}

res_tidy <- res |>
  as_tibble(rownames = "gene") |>
  select(gene, log2FoldChange, pvalue, padj)
```

Plotting the significant up/down regulated genes.

```{r}

genes <- c("AMBP", "ASCL1", "BARX1", "DIRAS3", "BNC1")

# Add to the plot

ggplot(res_tidy, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(aes(
    color = ifelse(
      !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 2, 
      "Significant", 
      "Not significant")
    ), 
    
    alpha = 0.8
) +
  labs(
    title = "Volcano plot for down/up regulated genes with significance < 0.05", 
    color = "Significant/ Unsignificant", 
    x = "log2FC",
    y = "-log10(pvalue)"
) +
  geom_point(
    data = res_tidy |> filter (gene %in% genes), 
    size=2,
    color = "black"
) +
  geom_label_repel(
    data = res_tidy |> filter (gene %in% genes), 
    aes(label=gene),
    size=2,
    color = "black", 
    max.overlaps=Inf,
    segment.color="transparent"
) +
  theme_minimal()

ggsave(
  here("results/05_volcano_plot.png"),
  plot= last_plot())
```

Extract significant differently express genes

```{r}
significant_genes <- res_tidy |>
  filter(padj<0.05 & abs(log2FoldChange)>2) |>
  select(gene)
```

Extract normalized counts

```{r}

normalized_counts <- counts(dds, normalized=TRUE)
```

```{r}

normalized_counts_flipped <- as.data.frame(normalized_counts) |>
  rownames_to_column("Gene") |>
  pivot_longer(cols = -Gene, names_to = "Sample", values_to = "Count") |>
  pivot_wider(names_from = "Gene", values_from = "Count") |>
  column_to_rownames(var="Sample")
```

Save Data

```{r}
write.csv(normalized_counts, here("data/05_normalized_counts.csv"))
write.csv(significant_genes, here("data/05_significant_genes.csv"))
```
